{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}

### 2018-01-09 thesis/background.tex
bgnd

{\bf Identifying bugs.}
Even if a program's behaviour is nondeterministic, that does not necessarily mean it has a bug.
After all, many programs use random number generation to intentionally generate different outputs.
We say a {\em concurrency bug} occurs when one or more of a program's nondeterministic behaviours is both {\em unanticipated} and {\em undesired}.
Most often, a concurrency novice who programs with shared state will consider the possible interleavings where one thread's access sequence occurs entirely before the other's, but neglect to consider intermediate outcomes in which the threads' access sequences are interleaved.
%
Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
	Exercise for the reader: Show why 2 is a possible output, but 1 is not!
},
but whether this constitutes a bug is a matter of perspective.
Was the program written to count to 2000, or was it written to compute a randomized distribution?
In this thesis, we make no attempt to reason about the ``intent'' of programs,
so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
according to commonly-accepted notions of what programs behaviours are always bad.
%
Bug conditions include assertion failures,
memory access errors (i.e., segmentation fault or bus error),
heap errors (i.e., use-after-free or overflow),
deadlocks,
and infinite loops (which must be identified heuristically \cite{entscheidungsproblem}).

@@

Transactional memory (TM) may be implemented either in hardware, using special instructions and existing cache coherence algorithms,
or in software, via library calls and a log-based commit approach.
Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose runtime overhead associated with logging.
Hardware transactions (HTM) \cite{htm-experience, htm-performance} achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
Haswell \cite{htm-haswell} is the first x86 architecture to support HTM,
offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
The example program in Figure~\ref{fig:htm-example} demonstrates how these instructions can be used to synchronize a simple shared access without locking overhead in the common case,
using GCC's compiler intrinsics \cite{htm-gcc}.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{l}
		\textbf{(All threads)} \\
		\hline
		\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
		\texttt{~~~~if (bad\_value\_of\_foo(foo)) \{} \\
		\texttt{~~~~~~~~\_xabort();} \\
		\texttt{~~~~\} else \{} \\
		\texttt{~~~~~~~~foo++;} \\
		\texttt{~~~~~~~~\_xend();} \\
		\texttt{~~~~\}} \\
		\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
		\texttt{~~~~mutex\_lock(\&m);} \\
		\texttt{~~~~foo++;} \\
		\texttt{~~~~mutex\_unlock(\&m);} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{An example program demonstrating the various HTM primitives.
		If the transaction in the top branch aborts,
		whether from a conflict or from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin},
		{\tt status} will be assigned an error code indicating the abort reason,
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

@@

@@

{\bf Reduction techniques.}
Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
Technically speaking, it identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.
Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
Figure~\ref{fig:dpor} shows a part of an execution tree in which the operations by threads 1 and 2 are independent, and hence only one of the subtrees need be tested.
Over the years, researchers have developed many enhancements to DPOR, such as Optimal DPOR \cite{optimal-dpor}, parallelizable DPOR \cite{parallel-dpor}, SAT-directed model checking \cite{satcheck}, Maximal Causality Reduction \cite{mcr}, and DPOR for relaxed memory architectures \cite{tsopso}.

>>>>

{\bf Identifying bugs.}
Even if a program's behaviour is nondeterministic, that does not necessarily mean it has a bug.
After all, many programs use random number generation to intentionally generate different outputs.
We say a {\em concurrency bug} occurs when one or more of a program's nondeterministic behaviours is both {\em unanticipated} and {\em undesired}.
Most often, a concurrency novice who programs with shared state will consider the possible interleavings where one thread's access sequence occurs entirely before the other's, but neglect to consider intermediate outcomes in which the threads' access sequences are interleaved.
%
Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
	Fun exercise for the reader: Show why 2 is a possible output, but 1 is not!
},
but whether this constitutes a bug is a matter of perspective.
Was the program written to count to 2000, or was it written to compute a randomized distribution?
In this thesis, we make no attempt to reason about the ``intent'' of programs,
so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
according to commonly-accepted notions of what programs behaviours are always bad.
%
Bug conditions include assertion failures,
memory access errors (i.e., segmentation fault or bus error),
heap errors (i.e., use-after-free or overflow),
deadlocks,
and infinite loops (which must be identified heuristically \cite{entscheidungsproblem}).

@@

Transactional memory (TM) may be implemented either in hardware, using special instructions and existing cache coherence algorithms,
or in software, via library calls and a log-based commit approach.
Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose runtime overhead associated with logging.
Hardware transactions (HTM) \cite{htm-experience, htm-performance} achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
Haswell \cite{htm-haswell} is the first x86 architecture to support HTM,
offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
The example program in Figure~\ref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
	The solution presented here is actually incomplete; stay tuned until Chapter~\ref{chap:tm} for the surprising twist!
},
using GCC's compiler intrinsics \cite{htm-gcc}.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{l}
		\texttt{void count() \{} \\
		\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		\texttt{~~~~~~~~if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\_xend();} \\
		\texttt{~~~~~~~~\} else \{} \\
		\texttt{~~~~~~~~~~~~mutex\_lock(\&m);} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~mutex\_unlock(\&m);} \\
		\texttt{~~~~~~~~\}} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
		If the transaction in the top branch aborts,
		whether from a memory conflict or random system interrupt,
		%from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin},
		{\tt status} will be assigned an error code indicating the abort reason,
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

@@

\subsection{Overview}

@@

\subsection{On State Space Explosion}

{\bf Reduction techniques.}
Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
Technically speaking, it identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.
More intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
% TODO: make a better fiugre based on the tree
Figure~\ref{fig:dpor} shows a part of an execution tree in which the operations by threads 1 and 2 are independent, and hence only one of the subtrees need be tested.
Over the years, researchers have developed many enhancements to DPOR, such as Optimal DPOR \cite{optimal-dpor}, parallelizable DPOR \cite{parallel-dpor}, SAT-directed model checking \cite{satcheck}, Maximal Causality Reduction \cite{mcr}, and DPOR for relaxed memory architectures \cite{tsopso}.
